Binning Strategies and Related Loss for Large Data

### 2015-05-12 Paper.Rnw
comments in section 5.2

\ktm{The reduced binned data from spatial binning contains the center and count information for all bins. The bin frequencies may be mapped continuously to a precisely rendered shade, however it is niave to believe that human perception will be able to perfectly extract that information. There is implicitly loss of frequency information occuring when the shade of a tile is visually mapped back to a frequency through the use of a shading scale index. Bin frequencies may themselves be binned in order to discretize the color scale for the binned scatterplot, thus making the loss explicit.

Figure~\ref{fig:StandardFreqBinning4710none} displays binned scatterplots with varying numbers of  standard binned frequency groups. If we attempt to discern differences between similarly shaded tiles: it is trivial when only four shades exist, it becomes more difficult -- but still possible -- at seven bins, and at ten frequency bins we are hardly able to discriminate better than in continuous shading. This aligns with Healey and Enn's theory on the number of discernible colors \citep{colorperception}. Our exploration of frequency loss will focus on frequency binning with at most ten bins because above this we experience implicit frequency from perceptual bounds that are not well reflected in the explicitly defined frequency loss. }

@@

The loss of frequency information in frequency binning is dependent on the selection of a discrete color mapping, where the binning algorithm and number of frequency bins must be specified. The top row in Figure~\ref{fig:FreqLossCompare} displays the frequency loss from using standard and quantile frequency binning algorithms, with between one and ten frequency bins, from each set of simulated data. We first notice that large difference in the magnitude of frequency losses based on the bivariate distribution; frequency losses are highest for the exponential data, lower for the normal data and lowest for the uniform data. We also note that the frequency loss is a decreasing function of the number of frequency bins for both binning algorithms. This relationship largely flattens out after four frequency bin, each subsequent bin reducing the frequency loss less than the previous. Thus we should consider using between four and seven bin shades in order to reduce loss while also allowing for easy perception of frequency groups in the binned scatterplot. Lastly, the frequency loss with the standard frequency binning is be higher than for the quantile-based algorithm when a minimal number of bins are used. But in all three simulated data sets, the losses drop to comparable levels for each algorithm when four or more frequency bins are used. \ktm{Therefore, it is strongly recommended to use the quantile-based algorithm to bin untransformed frequencies.}

>>>>

Spatial binning reduces data to contain the center and frequency information for all bins.
While bin frequencies could be mapped continuously to a precisely rendered shade, it is naive to believe that this information can be extracted at the same resolution due to limitations of our perception. 
Implicitly, we lose information  when (imperfectly) mapping the shade of a tile to its frequency. In order to make this loss explicit, we introduce a second stage of binning and discretize the color scale for frequencies in the spatially binned scatterplot.

Figure~\ref{fig:StandardFreqBinning4710none} displays several spatially binned scatterplots. On the left frequency is mapped continuously to colour; on the right  there is a varying numbers of  standard binned frequencies. It is trivial to discern differences between similarly shaded tiles
when there are only four shades. This becomes more difficult -- but still possible -- at seven frequency bins, but at ten frequency bins we are hardly able to discriminate between colours better than in the continuous shading. 
This aligns with Healey and Enn's theory on the number of discernible colors \citep{colorperception}. We will therefore use an upper limit of ten frequency bins in our exploration of frequency loss.

@@

The loss of frequency information in frequency binning is dependent on the selection of a discrete color mapping, where the binning algorithm and number of frequency bins must be specified. The top row in Figure~\ref{fig:FreqLossCompare} displays the frequency loss from using standard and quantile frequency binning algorithms, with between one and ten frequency bins, from each set of simulated data. 
\hh{XXX the next sentence needs to be fixed.}
\ktm{We first notice that large difference in the magnitude of frequency losses based on the bivariate distribution; frequency losses are highest for the exponential data, lower for the normal data and lowest for the uniform data.} We also note that the frequency loss is a decreasing function in the number of frequency bins for both binning algorithms. This relationship largely flattens out after four frequency bin, each subsequent bin reducing the frequency loss less than the previous. Thus we should consider using between four and seven bin shades in order to reduce loss while also allowing for easy perception of frequency groups in the binned scatterplot. Lastly, the frequency loss with the standard frequency binning is be higher than for the quantile-based algorithm when a minimal number of bins are used. But in all three simulated data sets, the losses drop to comparable levels for each algorithm when four or more frequency bins are used. \ktm{Therefore, it is strongly recommended to use the quantile-based algorithm to bin untransformed frequencies.}
